\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization \thanks{Matthew Gardner, Andrew McNabb, and Kevin Seppi are with the Department of Computer Science, Brigham Young University, 3361 TMCB, Provo, UT 84602 (phone: 801-422-8717; email: \{mjg82,a,k\}@cs.byu.edu).}

### 2010-04-08 honors_thesis/journal_paper/paper.tex
more brutality (I hope you like it)

For each case $\casegen \in \caseset$, we introduce an indicator function
$\indic_{t}^{\casegen}$.  This function is 1 for the realized case from
Table~\ref{tab:evals} and 0 for the other cases.  We can then sum over all of
the cases, and the indicator function will make all of the terms drop to zero
except for the case that actually occurs.  Although this update is usually
implemented using if-statements, it is convenient to describe it as a
mathematical expression.  For example, the indicator function for the specific
case $\casexn$ can be written as follows:

\begin{align}
  \nonumber
	\indic_{t+1}^{\casexn} & (\ofunc ( \ppos_{t} ) ,\ofunc(\leftn_{t}),
	\ofunc(\rightn_{t}) ,\ofunc(\pbest_{t-1}) ,\ofunc(\nbest_{t-1}))= \\
  \label{eq:deficasexn}
	&\begin{cases}
	   1 & \text{if} \ \ofunc(\ppos_{t}) < \ofunc(\pbest_{t-1}) \\
	   &\quad \wedge \quad \ofunc(\nbest_{t-1}) < \ofunc(\ppos_{t}) \\
	   &\quad \wedge \quad \ofunc(\nbest_{t-1}) < \ofunc(\leftn_{t}) \\
	   &\quad \wedge \quad \ofunc(\nbest_{t-1}) < \ofunc(\rightn_{t}) \\
	   0 & \text{otherwise}
	\end{cases}
\end{align}

For compactness in our equations we represent this without the parameters since
they can be inferred from the subscripts on $\indic_{t+1}^{\casexn}$.

We now provide notation for the velocity update equation.  Once again we use
$\casexn$ as an example:

\begin{align}
\nonumber
	\specvel_{t+1}^{\casexn} & (\pvel_t, \ppos_{t}, \leftn_{t}, \rightn_{t},
	\pbest_{t-1}, \nbest_{t-1}, \pURand_{t}, \nURand_{t}) \\
\nonumber
		&= \constriction \bigl[ \pvel_{t} +
			\pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
\nonumber
		& \quad \quad \quad \; + \ncoeff\nURand_{t}\otimes(\nbest_{t} -
		\ppos_{t}) \bigr] \\
\nonumber
		&= \constriction \bigl[ \pvel_{t} +
			\pcoeff\pURand_{t}\otimes(\ppos_{t} - \ppos_{t}) \\
\label{eq:defvcasexn}
			& \quad \quad \quad \; + \ncoeff\nURand_{t}\otimes(\nbest_{t-1} -
			\ppos_{t}) \bigr]
\end{align}

Note that we have substituted the actual values obtained for $\pbest$ and
$\nbest$ into Equation~\eqref{eq:defvcasexn}.  Since we have restricted
ourselves to a specific case, we know the values of $\pbest$ and $\nbest$ and
do not need those terms as parameters to $\specvel_{t+1}^{\casexn}$.  In this
case, $\pbest_{t}=\ppos_{t}$, as \pbest was updated by the particle's current
position, and $\nbest_{t}=\nbest_{t-1}$, as \nbest was not updated at iteration
$t$.  Again we will drop the parameters to $\specvel_{t+1}^{\casexn}$ for
compactness.

In the same way we can create notation for the position update
equation~\eqref{eq:velupdate}:
\begin{align}
\label{eq:defpcasexn}
	\specpos_{t+1}^{\casexn} & (\ppos_{t}, \pvel_{t}, \leftn_{t},
	\rightn_{t} ,\pbest_{t-1} ,\nbest_{t-1}, \pURand_{t}, \nURand_{t}) \\
\nonumber
	& = \ppos_{t} + \specvel_{t+1}^{\casexn}
\end{align}

With this notation we can re-write the original PSO velocity
equation~\eqref{eq:velupdate}, introducing our sum over cases with the
indicator functions.  The equation becomes:
\begin{align}
\nonumber
	\pvel_{t+1} &=
		\constriction \bigl[ \pvel_t
			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
\nonumber
			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} -
			\ppos_{t}) \bigr] \\
\nonumber
	&= \sum_{c \in \caseset} \indic_{t+1}^{c} \ \constriction \bigl[ \pvel_t
			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
\nonumber
			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} -
			\ppos_{t}) \bigr]  \\
\label{eq:vel2update}
	&= \sum_{c \in \caseset} \ \indic_{t+1}^{c} \ \specvel_{t+1}^{c} 
\end{align}

We can similarly re-write the position update equation~\eqref{eq:posupdate} as:
\begin{align}
\nonumber
	\ppos_{t+1} &= \ppos_{t} + \pvel_{t+1} \\
\label{eq:pos2update}
	&= \sum_{c \in \caseset} \ \indic_{t+1}^{c} \ \specpos_{t+1}^{c} 
\end{align}

@@

A particle at iteration $t-1$ that has been moved to iteration $t$ using
Equations~\eqref{eq:velupdate}~and~\eqref{eq:posupdate}, but whose position has
not yet been evaluated, is denoted as $\noeval{\p}_t$.  Once its position has
been evaluated, but it has still not yet received information from its
neighbors, it is denoted as $\nonbest{\p}_t$.  Only when the particle has
updated its neighborhood best is it a complete particle at iteration $t$.  It
is then simply denoted as $\p_t$.

@@

In a centralized, or Master-Slave, parallel PSO algorithm, one machine, the
master, keeps track of all necessary information, and all other machines are
merely used to evaluate the objective function at various positions as directed
by the master~\cite{belal-2004-parallel-models-for-pso}.  To perform speculative evaluation in such
an architecture, the master generates the positions to evaluate speculatively
as in Equation~\eqref{eq:pos2update}.  After having the slaves evaluate the
objective function at all necessary positions, the master then decides which
position to accept for each particle, as in Equation~\eqref{eq:val2updatelong}.
The outline of the procedure is given in \alg{centralized}.

\begin{algorithm}
  \caption{Speculative Evaluation in a Centralized PSO}
  \label{alg:centralized}
  \begin{algorithmic}[1]
	\STATE Move all $\p_{t-1}$ to $\noeval{\p}_t$ using
	  Equations~\eqref{eq:velupdate}~and~\eqref{eq:posupdate}
	\STATE For each $\noeval{\p}_t$, get its neighbors $\noeval{\nset}_t$ and
	  generate $\noeval{\sset}_{t+1}$ according to
	  Equation~\eqref{eq:pos2update}.
	\STATE Evaluate all $\noeval{\p}_t$ and $\noeval{\sset}_{t+1}$ in parallel
	\STATE Update personal best for each $\noeval{\p}_t$ and
	  $\noeval{\s}_{t+1}$, creating $\nonbest{\p}_t$ and $\nonbest{\s}_{t+1}$
	\STATE Update neighborhood best for each $\nonbest{\p}_t$, creating
	  $\pset_t$
	\FORALL{$\p_t$}
	\STATE Pick $\nonbest{\s}_{t+1}$ from $\nonbest{\sset}_{t+1}$ that matches
	  the branch taken by $\p_t$ according to
	  Equation~\eqref{eq:val2updatelong}.
	\STATE Pass along personal and neighborhood best values obtained by $\p_t$,
	  making $\nonbest{\p}_{t+1}$
	\ENDFOR
	\STATE Update neighborhood best for each $\nonbest{\p}_{t+1}$, creating
	  $\pset_{t+1}$
	\STATE Repeat from Step 1 until finished
  \end{algorithmic}
\end{algorithm}

Given a set of particles at iteration $t-1$ (perhaps which have just been
initialized), the master must move each particle using
Equations~\eqref{eq:velupdate} and \eqref{eq:posupdate} to obtain the set
$\noeval{\pset}_t$.  For each particle $\noeval{\p}_t$, the master must get its
set of neighbors $\noeval{\nset}_t$ and use their positions, along with the
position of $\noeval{\p}_t$, to enumerate all possible combinations of $\pbest$
and $\nbest$ that particle $\noeval{\p}_t$ could end up with (as shown in
Table~\ref{tab:evals}).  Each of those combinations defines a speculative
particle, $\s_t$, that can be moved using Equation~\eqref{eq:pos2update} to
obtain $\noeval{\s}_{t+1}$.  The result is a set of particles
$\noeval{\pset}_t$, and for each particle a set of speculative children
$\noeval{\sset}_{t+1}$, which can all be evaluated at once.

@@

This is done by first updating each particle's personal and neighborhood bests.
Mathematically, this corresponds to the evaluation of an indicator function
similar to that found in Equation~\eqref{eq:deficasexn}.  In practice, this is
done by moving each $\noeval{\p}_t$ to $\nonbest{\p}_t$ and each speculative
particle $\noeval{\s}_{t+1}$ to $\nonbest{\s}_{t+1}$ by associating the value
determined by the slave to the particle it came from, and updating that
particle's personal best if the value reported is better than its previous
personal best value.  Then each $\nonbest{\p}_t$ must be updated to $\p_t$ by
updating its neighborhood best position and value with information about each
of its neighbors.  This is simply the original PSO algorithm, and corresponds
to steps 1--5 in \alg{centralized}.

Then, with the set $\pset_t$, the branch that was actually taken by each
particle can be determined, and the correct speculative child can be chosen.
The branches depend entirely on whether or not the personal best of $p_t$ was
updated and which particle, if any, updated the particle's neighborhood best.
Both of those items are now known, and the child that matches the branch is
selected as per Equation~\eqref{eq:val2updatelong} (step 7 in
\alg{centralized}).

>>>>

For each case $\casegen \in \caseset$, we introduce an indicator function
$\indic_{t}^{\casegen}$.  This function is 1 for the realized case from
Table~\ref{tab:evals} and 0 for the other cases.  We can then sum over all of
the cases, and the indicator function will make all of the terms drop to zero
except for the case that actually occurs.  Although this update is usually
implemented using if-statements, it is convenient to describe it as a
mathematical expression.  For example, the indicator function for the specific
case $\casexn$ can be written as follows:
\begin{align}
  \nonumber
	\indic_{t+1}^{\casexn} & (\ofunc ( \ppos_{t} ) ,\ofunc(\leftn_{t}),
	\ofunc(\rightn_{t}) ,\ofunc(\pbest_{t-1}) ,\ofunc(\nbest_{t-1}))= \\
  \label{eq:deficasexn}
	&\begin{cases}
	   1 & \text{if} \ \ofunc(\ppos_{t}) < \ofunc(\pbest_{t-1}) \\
	   &\quad \wedge \quad \ofunc(\nbest_{t-1}) < \ofunc(\ppos_{t}) \\
	   &\quad \wedge \quad \ofunc(\nbest_{t-1}) < \ofunc(\leftn_{t}) \\
	   &\quad \wedge \quad \ofunc(\nbest_{t-1}) < \ofunc(\rightn_{t}) \\
	   0 & \text{otherwise}
	\end{cases}
\end{align}

For each case $\casegen \in \caseset$, there is also a corresponding
velocity update function $\specvel_{t+1}^{\casegen}$.  When the case is
known, the specific values of $\pbest_t$ and $\nbest_t$ may be substituted
directly into~\eqref{eq:velupdate}.  For example, in case $\casexn$,
$\pbest_{t}=\ppos_{t}$, as \pbest was updated by the particle's current
position, and $\nbest_{t}=\nbest_{t-1}$, as $\nbest$ was not updated at
iteration $t$:
\begin{align}
\nonumber
	\specvel_{t+1}^{\casexn} & (\pvel_t, \ppos_{t}, \leftn_{t}, \rightn_{t},
	\pbest_{t-1}, \nbest_{t-1}, \pURand_{t}, \nURand_{t}) \\
\nonumber
		&= \constriction \bigl[ \pvel_{t} +
			\pcoeff\pURand_{t}\otimes(\ppos_{t} - \ppos_{t}) \\
\label{eq:defvcasexn}
			& \quad \quad \quad \; + \ncoeff\nURand_{t}\otimes(\nbest_{t-1} -
			\ppos_{t}) \bigr]
\end{align}

In the same way we can create notation for the position update function by
substituting into~\eqref{eq:posupdate}.  For compactness, we will drop the
parameters to $\specvel_{t+1}^{\casegen}$ since they can be inferred from the
subscripts on $\indic_{t+1}^{\casegen}$.
\begin{align}
\label{eq:defpcasegen}
	\specpos_{t+1}^{\casegen} & (\ppos_{t}, \pvel_{t}, \leftn_{t},
	\rightn_{t} ,\pbest_{t-1} ,\nbest_{t-1}, \pURand_{t}, \nURand_{t}) \\
\nonumber
	& = \ppos_{t} + \specvel_{t+1}^{\casegen}
\end{align}

With this notation we can re-write the original PSO velocity
equation~\eqref{eq:velupdate}, introducing our sum over cases with the
indicator functions.  Again, we represent the indicator functions and velocity
functions without the parameters for compactness.  The equation becomes:
\begin{align}
\nonumber
	\pvel_{t+1} &=
		\constriction \bigl[ \pvel_t
			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
\nonumber
			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} -
			\ppos_{t}) \bigr] \\
\nonumber
	&= \sum_{c \in \caseset} \indic_{t+1}^{c} \ \constriction \bigl[ \pvel_t
			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
\nonumber
			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} -
			\ppos_{t}) \bigr]  \\
\label{eq:vel2update}
	&= \sum_{c \in \caseset} \ \indic_{t+1}^{c} \ \specvel_{t+1}^{c} 
\end{align}

Similarly, the position update equation~\eqref{eq:posupdate} becomes:
\begin{align}
\nonumber
	\ppos_{t+1} &= \ppos_{t} + \pvel_{t+1} \\
\label{eq:pos2update}
	&= \sum_{c \in \caseset} \ \indic_{t+1}^{c} \ \specpos_{t+1}^{c} 
\end{align}

@@

A particle at iteration $t-1$ that has been moved to iteration $t$ using
\eqref{eq:velupdate}~and~\eqref{eq:posupdate}, but whose position has
not yet been evaluated, is denoted as $\noeval{\p}_t$.  Once its position has
been evaluated, but it has still not yet received information from its
neighbors, it is denoted as $\nonbest{\p}_t$.  Only when the particle has
updated its neighborhood best is it a complete particle at iteration $t$.  It
is then simply denoted as $\p_t$.

@@

In a centralized, or Master-Slave, parallel PSO algorithm, one machine, the
master, keeps track of all necessary information, and all other machines are
merely used to evaluate the objective function at various positions as directed
by the master~\cite{belal-2004-parallel-models-for-pso}.  To perform speculative evaluation in such
an architecture, the master generates the positions to evaluate speculatively
as in \eqref{eq:pos2update}.  After having the slaves evaluate the
objective function at all necessary positions, the master then decides which
position to accept for each particle, as in \eqref{eq:val2updatelong}.
The outline of the procedure is given in \alg{centralized}.

\begin{algorithm}
  \caption{Speculative Evaluation in a Centralized PSO}
  \label{alg:centralized}
  \begin{algorithmic}[1]
	\STATE Move all $\p_{t-1}$ to $\noeval{\p}_t$ using
	  \eqref{eq:velupdate}~and~\eqref{eq:posupdate}
	\STATE For each $\noeval{\p}_t$, get its neighbors $\noeval{\nset}_t$ and
	  generate $\noeval{\sset}_{t+1}$ according to
	  \eqref{eq:pos2update}.
	\STATE Evaluate all $\noeval{\p}_t$ and $\noeval{\sset}_{t+1}$ in parallel
	\STATE Update personal best for each $\noeval{\p}_t$ and
	  $\noeval{\s}_{t+1}$, creating $\nonbest{\p}_t$ and $\nonbest{\s}_{t+1}$
	\STATE Update neighborhood best for each $\nonbest{\p}_t$, creating
	  $\pset_t$
	\FORALL{$\p_t$}
	\STATE Pick $\nonbest{\s}_{t+1}$ from $\nonbest{\sset}_{t+1}$ that matches
	  the branch taken by $\p_t$ according to
	  \eqref{eq:val2updatelong}.
	\STATE Pass along personal and neighborhood best values obtained by $\p_t$,
	  making $\nonbest{\p}_{t+1}$
	\ENDFOR
	\STATE Update neighborhood best for each $\nonbest{\p}_{t+1}$, creating
	  $\pset_{t+1}$
	\STATE Repeat from Step 1 until finished
  \end{algorithmic}
\end{algorithm}

Given a set of particles at iteration $t-1$ (perhaps which have just been
initialized), the master must move each particle using
\eqref{eq:velupdate} and \eqref{eq:posupdate} to obtain the set
$\noeval{\pset}_t$.  For each particle $\noeval{\p}_t$, the master must get its
set of neighbors $\noeval{\nset}_t$ and use their positions, along with the
position of $\noeval{\p}_t$, to enumerate all possible combinations of $\pbest$
and $\nbest$ that particle $\noeval{\p}_t$ could end up with (as shown in
Table~\ref{tab:evals}).  Each of those combinations defines a speculative
particle, $\s_t$, that can be moved using \eqref{eq:pos2update} to
obtain $\noeval{\s}_{t+1}$.  The result is a set of particles
$\noeval{\pset}_t$, and for each particle a set of speculative children
$\noeval{\sset}_{t+1}$, which can all be evaluated at once.

@@

This is done by first updating each particle's personal and neighborhood bests.
Mathematically, this corresponds to the evaluation of an indicator function
similar to that found in \eqref{eq:deficasexn}.  In practice, this is
done by moving each $\noeval{\p}_t$ to $\nonbest{\p}_t$ and each speculative
particle $\noeval{\s}_{t+1}$ to $\nonbest{\s}_{t+1}$ by associating the value
determined by the slave to the particle it came from, and updating that
particle's personal best if the value reported is better than its previous
personal best value.  Then each $\nonbest{\p}_t$ must be updated to $\p_t$ by
updating its neighborhood best position and value with information about each
of its neighbors.  This is simply the original PSO algorithm, and corresponds
to steps 1--5 in \alg{centralized}.

Then, with the set $\pset_t$, the branch that was actually taken by each
particle can be determined, and the correct speculative child can be chosen.
The branches depend entirely on whether or not the personal best of $p_t$ was
updated and which particle, if any, updated the particle's neighborhood best.
Both of those items are now known, and the child that matches the branch is
selected as per \eqref{eq:val2updatelong} (step 7 in
\alg{centralized}).
